In this thesis, we formally verify a proof of the Cook-Levin Theorem in the proof assistant Coq.
The Cook-Levin Theorem states that the satisfiability problem SAT of conjunctive normal forms is NP-complete. 
This means that any language which is polynomial-time verifiable is reducible to SAT in polynomial time. 
Despite its importance, most proofs do not even attempt to verify the construction's correctness.

%The theorem was proved in 1971 by Stephen A. Cook~\cite{cook_theorem} and was independently discovered by Leonid Levin in 1973~\cite{levin_theorem}.
%Since the subsequent discovery of more NP-complete problems by Richard Karp in 1972~\cite{Karp1972}, it has become one of the most important results of computational complexity theory.

For the proof, one has to encode computations of a chosen reasonable computational model using Boolean formulas. 
We use the call-by-value $\lambda$-calculus L as the computational model as part of a larger effort to formalise the basics of complexity theory. 
%which has been shown to be polynomial-time and linear-space equivalent to Turing machines~\cite{ForsterKunzeRoth:2019:wcbv-Reasonable}. 
It is much easier to write and verify L-programs than to construct Turing machines.
%, which is why L is the model of choice for a larger effort to formalise the basics of complexity theory.

Thus, we need to reduce the computation of L-terms to SAT\@.
As Turing machines are structurally simpler than the $\lambda$-calculus, we propose to use Turing machines as an intermediate problem in the reduction of L to SAT, performing the computation of a Turing machine from an L-term and the construction of a SAT-formula from a Turing machine in L.

In this work, we introduce the basics of formalising polynomial-time reductions in Coq. We present a polynomial-time reduction from Turing machines to SAT in L which is based on the original tableau construction by Cook and formally verify its correctness and resource usage. The original construction is adapted and factorised in order to make a formal proof feasible. We see this as a significant first step towards showing SAT to be NP-complete in L, leaving the reduction from L to Turing machines for future work.
